The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration

### 2022-09-05 manuscript/manuscript.Rmd
methods

**Model structure.** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (`growth curve` or `Richard's curve`) of the form:

@@

Similarly, $\mu_j$ is inferred for locations with ecological proxies from the associated normal temperature distributions with a given mean and standard deviation, $mean_j$ and $sd_j$, as

$$mean_j  \sim N(\, \mu_j,sd_j).  $$

This structure implies that $\mu_j$ is not fixed at the mean proxy temperature at location $j$, <!--of the temperatures $t_{i=1,...,m,~j}$ of location $j$-->, but is drawn (`regularised`) towards the overall logistic regression curve. The pull towards the regression curve tends to be high when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, and/or when the overall standard deviation $\sigma$ is low. In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the overall regression than well-sampled locations with consistent temperature observations.

**Priors.** In the Bayesian framework, priors need to be placed on the unknown parameters of a model. We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$:
$$\sigma  \sim \sqrt{Inv{\text -}Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $$
$$\sigma_j  \sim \sqrt{Inv{\text -}Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n. $$
We set $\alpha = 1$ and $\beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, which is desirable as we have little a-priori knowledge of these parameters.

In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles and knowledge from the modern climate system:
1) Predicted seawater surface temperatures is not allowed to be $<< -2^\circ C$, the freezing point of sea water.
2) Input of solar energy decreases from the tropics to the poles. Hence, the latitudinal temperature gradient is generally negative, i.e. temperature decreases with absolute latitude. This is achieved by constraining $K-A$ to be positive.
3) The steepness of the gradient is likely highest in mid-latitudes.
4) The 
a skew-normal prior on the lower asymptote $A$, with ...  . This constrains $A$ to not be $<< -2^\circ C$, which is the freezing point of sea water. We chose this distribution such that its mode is $~ 0^\circ C$, slightly higher than the modern...

- logistic regression
- hierarchical model structure
- choice of priors (based on modern, see below)

>>>>

**Model structure (Fig 2).** We model the mean temperature ($\mu$) at location $j$ as a function of absolute latitude ($abs(l)$) with a logistic regression (`growth curve` or `Richard's curve`) of the form:

@@

Similarly, $\mu_j$ is inferred for locations with ecological proxies from the associated normal temperature distributions with a given mean and standard deviation, $t_{\mu,j}$ and $t_{\sigma,j}$, as

$$t_{\mu,j}  \sim N(\mu_j,t_{\sigma,j}).  $$

This structure implies that $\mu_j$ is not fixed at the mean proxy temperature at location $j$, <!--of the temperatures $t_{i=1,...,m,~j}$ of location $j$-->, but is drawn towards the overall logistic regression curve, i.e. towards $\nu_j$. The pull towards $\nu_j$ tends to be high when $m$ is low, when the observations $t_{i=1,...,m,~j}$ are scattered, i.e. $\sigma_j$ is high, and/or when the overall standard deviation $\sigma$ is low. In practice, this has the desirable consequence that locations with few observations and large temperature differences between observations have less influence on the overall regression than well-sampled locations with consistent temperature observations.

**Priors (Fig2x).** In the Bayesian framework, priors need to be placed on the unknown parameters of a model. We placed weakly informative, conjugate inverse-gamma priors on $\sigma$ and $\sigma_{j=1,..n}$:
$$\sigma  \sim \sqrt{Inv-Gamma(\alpha+\frac{n}{2},\beta + 0.5\times(\mu_j-\nu_j))}, ~~~~~ j = 1,...,n,  $$
$$\sigma_j  \sim \sqrt{Inv-Gamma(\alpha+\frac{m}{2},\beta + 0.5\times(t_{i,j}-\mu_j))}, ~~~~~ i = 1,...,m,~~~~~j = 1,...,n. $$
We set $\alpha = \beta = 1$, allowing these priors to be quickly overwhelmed by the data as $n$ and $m$ increase, as we have little *a priori* knowledge of these parameters.

In contrast, we put informative priors on the regression coefficients $A$, $K$, $M$ and $B$, based on physical principles, and vaguely based on the modern climate system:

**A.** Predicted seawater surface temperatures are not allowed to be $<< -2^\circ C$, the freezing point of sea water. The highest prior density of $A$ is placed around $0^\circ C$, and it slowly tapers off towards higher temperatures. This shape is achieved by placing a skew-normal prior on the lower asymptote, specified as 
$$A \sim SN(\xi=-3.0,\omega=12,\alpha_{SN}=30),$$
where $\xi$, $\omega$, and $\alpha_{SN}$ are the location, scale and shape parameters.

**K.** Input of solar energy decreases from the tropics to the poles. Hence, the latitudinal temperature gradient is broadly negative, i.e. temperature decreases with absolute latitude. This is achieved by setting $K \ge A$. The prior on the upper asymptote $K$ is a truncated normal distribution with the mean set to $K$ of the modern SST gradient, with a broad standard deviation: 
$$K \sim TN(\mu_{TN} = 28, \sigma_{TN} = 10, \alpha_{TN} = A, \beta_{TN}= \infty)$$
The distribution is truncated to the left at $\alpha_{TN}=A$, but not truncated to the right ($\beta_{TN}$).

**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to $M$ of the modern SST gradient: 
$$M \sim N(42,10)$$
**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to prevent very abrupt SST changes across latitudes. A gamma-distributed prior of the form
$$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$$
was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$.
